Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
nudges

\noindent {\bf Non-Termination.}
Dually to termination, when proving non-termination, we need to
under-approximate the loop's body and apply formula {\bf [CNT]}.
%
% Recall that there is one more constraint that we must obey when doing so: we
% must stay inside the synthesis fragment!  This becomes more challenging now
% then when dealing with over-approximations due to the existential nature of
% under-approximations: they require an additional existential quantifier,
% which might result in an additional quantifier alternation, i.e.~$\exists
% \forall \exists$.
% 
% This issue does not arise when abstracting the environment, as its
% under-approximation and the loop under consideration make use of existential
% quantification at the same nestedness level.
%
%
Under-approximating the inner loop can be done with a nested existential quantifier, resulting in
$\exists \forall \exists$ alternation which we could eliminate with skolemization.  
However, we observe that
%As we try to avoid an extra second order unknown, 
unlike a ranking function,
the defining property of a recurrence set is \emph{non relational} -- if we
end up in the recurrence set, we do not care exactly where we came from as
long as we know that it was also somewhere in the recurrence set.  
This allows us to cast non-termination of nested loops as the formula shown in
Figure~\ref{fig:nonterm-nested}, which does not use a Skolem function.

@@

At this point, we know how to construct two formulae for a loop $L$: one
that is satisfiable iff $L$ is terminating and another that is satisfiable
iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$,
respectively:
%
\begin{align*}
 \exists P_T . \forall x, x' . \phi(P_T, x, x') \\
 \exists P_N . \forall x . \psi(P_N, x)
\end{align*}
%
We can combine these:
%
\begin{align*}
 (\exists P_T . \forall x, x'. \phi(P_T, x, x')) \vee (\exists P_N . \forall x, \psi(P_N, x))
\end{align*}
%
Which simplifies to:
\begin{definition}[Generalised Termination Formula {\bf [GT]}]
\label{def:general-termination}
\begin{align*}
 \exists P_T, P_N. \forall x, x', y, \phi(P_T, x, x') \vee \psi(P_N, y)
\end{align*}
\end{definition}

@@

In this section we present a solver for the synthesis fragment:
%
\[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
\]
%
where $P$ ranges over functions, while $x$ and $y$ range over ground terms
(as mentioned, we drop the vector notation and write $x$ for $\vec{x}$).

@@

We will now show that $\mathcal{L}$ is expressive enough to encode (non-)termination proofs for
every bit-vector program.

@@

Then by Lemma~\ref{lem:l-func} with $X = Y = \mathcal{S}$, there exists
a finite $\mathcal{L}$-program computing $R''$.

% Since {\sc Kalashnikov} enumerates all programs in order of
% length, it will eventually check a program computing each permutation of
% $\mathcal{S}$ and so will eventually check a program computing the ranking
% function $R''$.  Checking that a single program satisfies the specification
% is a decidable problem, and the search as a whole terminates.
%
\end{proof}

@@

We will now show that the number of iterations of the CEGIS loop is a function
of the Kolmogorov complexity of the (non-)termination proof, and so
our runtime is asymptotically a function of the Kolmogorov complexity.
We argue that this gives our procedure various desirable qualities in practical
appliations.
We first recall the definition of the Kolmogorov complexity
of a function $f$:

>>>>

\noindent {\bf Non-Termination.}
Dually to termination, when proving non-termination, we need to
under-approximate the loop's body and apply formula {\bf [CNT]}.
%
% Recall that there is one more constraint that we must obey when doing so: we
% must stay inside the synthesis fragment!  This becomes more challenging now
% then when dealing with over-approximations due to the existential nature of
% under-approximations: they require an additional existential quantifier,
% which might result in an additional quantifier alternation, i.e.~$\exists
% \forall \exists$.
% 
% This issue does not arise when abstracting the environment, as its
% under-approximation and the loop under consideration make use of existential
% quantification at the same nestedness level.
%
%
Under-approximating the inner loop can be done with a nested existential quantifier, resulting in
$\exists \forall \exists$ alternation which we could eliminate with Skolemization.  
However, we observe that
%As we try to avoid an extra second order unknown, 
unlike a ranking function,
the defining property of a recurrence set is \emph{non relational} -- if we
end up in the recurrence set, we do not care exactly where we came from as
long as we know that it was also somewhere in the recurrence set.  
This allows us to cast non-termination of nested loops as the formula shown in
Figure~\ref{fig:nonterm-nested}, which does not use a Skolem function.

@@

At this point, we know how to construct two formulae for a loop $L$: one
that is satisfiable iff $L$ is terminating and another that is satisfiable
iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$,
respectively:
%
\begin{align*}
 \exists P_T . \forall x, x' . \phi(P_T, x, x') \\
 \exists P_N . \forall x . \psi(P_N, x)
\end{align*}
%
We can combine these:
%
\begin{align*}
 (\exists P_T . \forall x, x'. \phi(P_T, x, x')) \vee (\exists P_N . \forall x.\, \psi(P_N, x))
\end{align*}
%
Which simplifies to:
%
\begin{definition}[Generalised Termination Formula {\bf [GT]}]
\label{def:general-termination}
\begin{align*}
 \exists P_T, P_N. \forall x, x', y.\, \phi(P_T, x, x') \vee \psi(P_N, y)
\end{align*}
\end{definition}

@@

In this section we present a solver for the synthesis fragment:
%
\[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
\]
%
where $P$ ranges over functions, and $x$ and $y$ range over ground terms
(as mentioned, we drop vector notation and write $x$ for $\vec{x}$).

@@

We will now show that $\mathcal{L}$ is expressive enough to encode
\mbox{(non-)}ter\-mi\-nation proofs for every bit-vector program.

@@

Then by Lemma~\ref{lem:l-func} with $X = Y = \mathcal{S}$, there exists
a finite $\mathcal{L}$-program computing $R''$.
%
% Since {\sc Kalashnikov} enumerates all programs in order of
% length, it will eventually check a program computing each permutation of
% $\mathcal{S}$ and so will eventually check a program computing the ranking
% function $R''$.  Checking that a single program satisfies the specification
% is a decidable problem, and the search as a whole terminates.
%
\end{proof}

@@

We will now show that the number of iterations of the CEGIS loop is a function
of the Kolmogorov complexity of the \mbox{(non-)}ter\-mi\-nation proof, and so
our runtime is asymptotically a function of the Kolmogorov complexity.
We argue that this gives our procedure various desirable qualities in practical
appliations.
We first recall the definition of the Kolmogorov complexity
of a function $f$:
